---
title: "Geography 13"
author: "[Piper Lovegreen] (https://piper-lovegreen.github.io/)"
subtitle: 'Lab 03: Distances and the Border Zone'
output:
  html_document:
    theme: journal
---
```{r}
# SPDS
library(tidyverse)
library(sf)
library(units)
library(readxl)

# Data
library(USAboundaries)
library(rnaturalearth)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)
```

# Question 1 
## 1.1 Define a Projection- North America Equidistant Conic
### Equidistant preserves distance
```{r}
eqdc = '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```
## 1.2 Get USA State Boundaries 
```{r}
conus= USAboundaries::us_states(resolution = "low") %>% 
  filter(!name %in% c("Hawaii", "Alaska", "Puerto Rico")) %>% 
  st_transform(eqdc)
```
## 1.3 Get County Boundaries for Mexico, USA, and Canada
```{r}
countries= rnaturalearth::countries110 %>% 
  st_as_sf(countries) %>%  
  filter(admin %in% c("United States of America", "Mexico", "Canada")) %>% 
  st_transform(eqdc)
```
## 1.4 Get City Locations from the CSV File 
```{r}

url = 'https://raw.githubusercontent.com/mikejohnson51/exercise-04/master/data/uscities.csv'
us_cities = readr::read_csv(url) %>% 
  filter(!state_name %in% c("Hawaii", "Puerto Rico", "Alaska")) %>% 
  st_as_sf(coords= c("lng", "lat"), crs= 4326) %>% 
  st_transform(eqdc)

```

# Question 2 
## 2.1 Distance to USA Border (Coastline or National) [km]
```{r}
conus = USAboundaries::us_states() %>%
  filter(!state_name %in% c("Puerto Rico",
                            "Alaska",
                            "Hawaii")) 
conus <- st_transform(conus, eqdc)

us_border = st_union(conus) %>%
    st_cast("MULTILINESTRING")
plot(us_border)

border_distances= us_cities %>% 
  mutate(distance_border= st_distance(., us_border),
  distance_border= units::set_units(distance_border, "km"),
  distance_border= as.numeric(distance_border))

table2.1= border_distances %>% 
  select("city", "state_name", "distance_border") %>% 
  slice_max(distance_border, n= 10)  

  
knitr::kable(table2.1,
              caption= "10 Furthest Cities from USA Border",
              col.names= c( "City", "State Name", "Distance to Border [km]", "Longitude and Latitude"),
             format.args = list(big.mark = ",")) %>% 
              
kableExtra::kable_styling("basic", full_width = TRUE, font_size = 14)
  
```
## 2.2 Distance to States [km]
```{r}
conus = USAboundaries::us_states() %>%
  filter(!state_name %in% c("Puerto Rico",
                            "Alaska",
                            "Hawaii")) 
conus <- st_transform(conus, eqdc)


state_border=  st_combine(conus) %>%
    st_cast("MULTILINESTRING")
plot(state_border)

border_distances_state= us_cities %>% 
  mutate(statedistance_border= st_distance(., state_border),
  statedistance_border= units::set_units(statedistance_border, "km"),
  statedistance_border= as.numeric(statedistance_border))

table2.2= border_distances_state %>% 
  select( "city", "state_name","statedistance_border") %>% 
  slice_max(statedistance_border, n= 10)

knitr::kable(table2.2,
              caption= "Distance from Cities to State Borders",
              col.names= c("City", "State Name", "Distance to Border[km]", "Longitude and Latitude"),
              format.args= list(big.mark= ".")) %>% 
  kableExtra::kable_styling("basic", full_width = TRUE, font_size = 14)
```
## 2.3 Distance to Mexico [km]
```{r}
mex= countries %>% 
  filter(admin %in% 'Mexico')
  
 mex_border= st_union(mex) %>%
    st_cast("MULTILINESTRING")
plot(mex_border)

mex_distance= us_cities %>% 
  mutate(mex_dis_border=st_distance(., mex),
         mex_dis_border=units::set_units(mex_dis_border, "km"),
  mex_dis_border= as.numeric(mex_dis_border))

table2.3= mex_distance %>% 
  select("city", "state_name", "mex_dis_border") %>% 
  slice_max(mex_dis_border, n=10)

knitr::kable(table2.3,
              caption= "Top 10 Furthest Cities from Mexico Border",
              col.names= c("City", "State Name",  "Distance to Border[km]", "Longitude and Latitude"),
              format.args= list(big.mark= ".")) %>% 
kableExtra::kable_styling("basic", full_width = TRUE, font_size = 14)
       

```
## 2.4 Distance to Canada [km]
```{r}
can= countries %>% 
  filter(admin %in% 'Canada')
(can_border = st_union(can) %>%
    st_cast("MULTILINESTRING"))
plot(can_border)

can_distance= us_cities %>% 
  mutate(can_dis_border=st_distance(., can_border),
         can_dis_border=units::set_units(can_dis_border, "km"),
  can_dis_border= as.numeric(can_dis_border))

table2.4= can_distance %>% 
  select("city","state_name","can_dis_border") %>% 
  slice_max(can_dis_border, n=10)

knitr::kable(table2.4,
              caption= "Top 10 Furthest Cities from Canada Border",
              col.names= c("City", "State Name", "Distance to Border[km]", "Longitude and Latitude"),
              format.args= list(big.mark= ".")) %>% 
kableExtra::kable_styling("basic", full_width = TRUE, font_size = 14)


```
# Question 3 
## 3.1 Data
```{r}
borders = st_combine(countries) %>%
    st_cast("MULTILINESTRING")
plot(borders)

state_border=  st_combine(conus) %>%
    st_cast("MULTILINESTRING")
plot(state_border)

max_pop= us_cities %>% 
  slice_max(population, n= 10)

ggplot() + 
  geom_sf(data = borders, aes()) + 
  geom_sf(data=state_border, aes())+
  geom_sf(data= max_pop, aes(), col="coral1") + 
  ggrepel::geom_label_repel(
    data = max_pop,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  labs(title = "10 Cities with Largest Populations",
       subtitle =" In the United States") + 
  ggthemes::theme_map() 
```
## 3.2 City Distance from the Border
```{r}
borders = st_combine(countries) %>%
    st_cast("MULTILINESTRING")
plot(borders)

state_border=  st_combine(conus) %>%
    st_cast("MULTILINESTRING")
plot(state_border)

border_distances= us_cities %>% 
  mutate(distance_border= st_distance(., us_border),
  distance_border= units::set_units(distance_border, "km"),
  distance_border= as.numeric(distance_border))

top_5= border_distances %>% 
  slice_max(distance_border, n=5)

ggplot() + 
  geom_sf(data = border_distances, aes(col= distance_border)) +
  scale_color_gradient(low= "yellowgreen", high= "red1")+
  geom_sf(data=top_5, aes(), col= "red4")+
  geom_sf(data = us_border, aes()) + 
  ggrepel::geom_label_repel(
    data = top_5,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  labs(title = "5 Cities Furthest from US National Border",
       col= "Distance to Border [km]") + 
  ggthemes::theme_map()+
theme(legend.position = "right")
  
```
## 3.3 City Distnace from Nearest State
```{r}
state_border=  st_combine(conus) %>%
    st_cast("MULTILINESTRING")
plot(state_border)

border_distances_state= us_cities %>% 
  mutate(statedistance_border= st_distance(., state_border),
  statedistance_border= units::set_units(statedistance_border, "km"),
  statedistance_border= as.numeric(statedistance_border))
ggplot() + 
  geom_sf(data = border_distances_state, aes(col= statedistance_border)) +
  scale_color_gradient(low= "yellowgreen", high= "red1")+
  geom_sf(data=top_5, aes(), col= "red4")+
  geom_sf(data = state_border, aes()) + 
  ggrepel::geom_label_repel(
    data = top_5,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  labs(title = "City Distance from State Border",
         subtitle= "5 Cities Furthest from US National Border",
       col= "Distance to State Border [km]") + 
  ggthemes::theme_map()+
theme(legend.position = "right")
```
## Equidistance Boundary from Mexico and Canada
```{r}
city3.4= us_cities %>% 
  select(city, state_name, population, geometry)

uscities3.4= city3.4 %>% 
  filter(!state_name %in% c("Alaska", "Hawaii", "Puerto Rico")) %>% 
  st_as_sf(coords= c("lng", "lat"), crs= 4326) %>% 
    st_transform(eqdc)
  

mex_can_dis= uscities3.4 %>% 
  mutate(can_dis= st_distance(uscities3.4, can_border),
         can_dis= units::set_units(can_dis, "km"),
         can_dis= as.numeric(can_dis)) %>% 
  ungroup() %>% 
  mutate(mex_dis= st_distance(uscities3.4, mex_border),
         mex_dis= units::set_units(mex_dis, "km"),
         mex_dis= as.numeric(mex_dis)) %>% 
  ungroup()

ab= mex_can_dis %>%   
  mutate(ab_dis= abs(can_dis - mex_dis)) %>% 
  select(city, state_name, population, can_dis, mex_dis, ab_dis)

pop_3.4 <- ab %>% 
  filter(ab_dis <= 100) %>% 
  slice_max(population, n=5)


ggplot() + 
  geom_sf(data = borders, aes()) + 
  geom_sf(data=state_border, aes())+
  geom_sf(data= ab, aes(col= ab_dis)) +
  gghighlight::gghighlight(ab_dis <= 100)+
geom_sf(data= pop_3.4, aes(), col= "coral1") +
  ggrepel::geom_label_repel(
    data = pop_3.4,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  labs(title = "US Cities 100km from Mexico and Canada Border",
       subtitle ="5 Most Populous Cities within this Zone") + 
  ggthemes::theme_map() 
```
# Question 4 
## 4.1 Quantifing Border Zone 
```{r}
border_distances= us_cities %>% 
  mutate(distance_border= st_distance(., us_border),
  distance_border= units::set_units(distance_border, "km"),
  distance_border= as.numeric(distance_border)) %>% 
  filter(distance_border <= 160) %>% 
  select(city, state_name, population, distance_border, geometry)  
  
 
city_100zone= nrow(border_distances)
pop_tot= sum(border_distances$population)
pop_percent= (pop_tot/ sum(us_cities$population) *100) %>% 
  round(digits=2)
 
table4.1= data.frame(Description= c("Number of Cities within 100mi Zone", "Total Population within 100mi Zone", "Percent of Continental US within 100mi Zone"), Numbers= c(city_100zone, pop_tot, pop_percent))
knitr::kable(table4.1,
              caption= "Information about 100 mile Zone of US Border") %>% 
kableExtra::kable_styling("basic", full_width = TRUE, font_size = 14)


```
### 4.2 Mapping Border Zone 
```{r}
city_pop= border_distances %>% 
  filter(distance_border <= 160) %>% 
  group_by(state_name) %>% 
  slice_max(population, n= 10)

buffer100= border_distances %>% 
  filter(distance_border <= 160) 



ggplot()+
  geom_sf(data= state_border, aes())+
    geom_sf(data= buffer100, aes(col= distance_border))+
  scale_color_gradient(low= "orange", high= "darkred")+
  ggrepel::geom_label_repel(
    data = city_pop,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  labs(title = "Top 10 Most Populous Cities in 100mile Zone",
       col= "Distance from Border") + 
  ggthemes::theme_map()+
theme(legend.position = "right")
```





